---
title: "Assignment 2: Coding Basics"
author: "Keanu Valibia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 

# Generate a sequence of numbers, 1 - 30, skipping every 3 numbers
genseq <- seq(1,30,3)
genseq

#2. 

# Use the mean() function to generate mean of my genseq object. 
#Store this value as a separate object.
mean_genseq <- mean(genseq) 

# Show mean value
mean_genseq 

# Use the median() function to generate median of my genseq object. 
#Store this value as a separate object.
med_genseq <- median(genseq) 

# Show median value
med_genseq 

#3. 

# Compare mean_genseq and med_genseq. Prints TRUE or FALSE value.
mean_genseq > med_genseq
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

# Create vector that holds student names
studentnames <- c("Keanu","Jay","Camila","Chris")
# Print vector
studentnames

# Create vector that holds test scores
test_scores <- c(100,25,50,49)
# Print vector
test_scores

# Create vector that holds logical values
test_results <- c(TRUE,FALSE,TRUE,FALSE)
# Print vector
test_results

# Create data frame, combining above three vectors
df_student_test_results <- data.frame("StudentName" = studentnames,
  "TestScore" = test_scores, 
  "TestResult" = test_results) 

# Call data frame I just created
df_student_test_results
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame allows for multiple classes to be held in the same object, whereas a matrix only allows usage of a single type of class.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}

# Create function using 'ifelse' logic
resultcheck <- function(result) {
  # If 'result < 50' is TRUE, print "FALSE...."
  print(ifelse(result < 50, FALSE, TRUE))  # ...otherwise, print "TRUE"
}

for (i in length(test_scores)) { # For every value in 'test_scores' vector...
  resultcheck(test_scores) #...applies 'resultcheck' function and print results
}

resultcheck2 <- function(result) { 
  if (result < 50) { # If logic check: if result is less than 50...
    print(FALSE) # ..print FALSE
  }
  else { # else: if result is 50 or larger...
    print(TRUE) # ...print TRUE
  }
}

#  for (i in length(test_scores)) { For every value in 'test_scores' vector...
#   resultcheck2(test_scores) Applies 'resultcheck' function and print results
# }
# 
# But this fails because it only tests for the first component in the vector

# Below logic resolves the error by explicitly asking to test individual 
# components in 'test_scores' until it reaches end of length
for (i in 1:length(test_scores)) { # For every component in 'test_scores'...
  resultcheck2(test_scores[i]) # Applies 'resultcheck' and print results
}
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The 'ifelse' statement worked. The 'if' logic seems to only check for the first value in the vector, and returns an error message beyond the first component. 'ifelse' is a vectorized function, and so is able to automatically test each component within a vector.