---
title: "Assignment 3: Data Exploration"
author: "Keanu Valibia"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r set.up, message=FALSE}

library(tidyverse) # Load tidyverse package
library(lubridate) # Load lubridate package

getwd() # Retrieve current working directory

# Create Neonics object using ecotox .csv
Neonics <- read.csv("~/R/R Projects/EDA_Spring2024/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", 
  stringsAsFactors = TRUE)

#Create Litter object using NEON .csv
Litter <- read.csv("~/R/R Projects/EDA_Spring2024/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", 
  stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We might be interested in understanding the specific chemical effects on insects, which may have drastic effects on their populations 
and surrounding environments. Chemical use may also have unpredictable effects, including changes in behavior or lifespan.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Forest litter and debris is a source of energy for forests and aquatic systems. It also provides shelter for smaller animals / insects.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. (Spatial sampling) Sampling is executed at terrestrial NEON sites that contain woody vegetatioon less than 2 meters tall.
 2. (Spatial sampling) In sites with forested tower airsheds, the litter sampling is targeted to take place in 20 40m x 40m plots
 3. (Temporal sampling) Frequent sampling (1x every 2 weeks) in deciduous forest sites &  infrequent sampling (1x every 1-2 months) 
 at evergreen sites

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r dimensions}
# Use dim() function to retrieve dimensions of Neonics dataset
dim(Neonics)
```

> Answer: 4,623 Rows and 30 Fields/Columns. These same dimensions can be viewed in the "Environment" panel.

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r neonics.effect.summary}
# Create vector object containing summary() of Effect field from Neonics
neonics.effect.summary <- summary(Neonics$Effect)
# Call newly created object
neonics.effect.summary

# Sort summary by descending order and print results
neonics.effect.summary[order(neonics.effect.summary, decreasing = TRUE)]
```

> Answer: The most common effects are Population and Mortality. We are interested in these effects to see:
1. Population-level effects as a result of chemical use and
2. Effects of insect mortality due to chemical use

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r neonics.species.summary}
# Create vector object containing summary() of Species.Common.Name field from Neonics
neonics.species.summary <- sort(summary(Neonics$Species.Common.Name), decreasing = TRUE)
# Call newly created object
neonics.species.summary
```

> Answer: Excluding (Other), the top six include:
1. Honeybee
2. Parasitic Wasp
3. Buff Tailed Bumblebee
4. Carniolan Honey Bee
5. Bumble Bee
6. Italian Honeybee
We may be interested particularly in bees due to their prominent role in pollination.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r author.class}
# Check field class using class() function
class(Neonics$Conc.1..Author.)
```

> Answer: It is a factor. It is not numeric because it includes non-numeric values. Thus RStudio defaults this field class to the most inclusive class.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r publication.plot}
# Create new object to save plot
freqPlot.publications1 <- ggplot(Neonics, aes(x = Publication.Year)) +
  # Frequency plot type
  geom_freqpoly() + 
  # Change theme
  theme_minimal() + 
  # Create title and subtitle
  ggtitle("Number of Publications per Year", 
          "Source: National Ecological Observatory Network") +
  # Change label names
  labs(x = "Year", y = "Number of Publications")

# Call frequency plot
freqPlot.publications1
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r publication.plot2}
freqPlot.publications2 <- ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  # Frequency plot type
  geom_freqpoly() +
  # Change plot theme
  theme_minimal() +
  # Set title and subtitle
  ggtitle("Number of Publications per Year", 
          "Source: National Ecological Observatory Network") +
  # Set x- and y-axis lables. Change legend title.
  labs(x = "Year", y = "Number of Publications", color = "Test Location")

# Call frequency plot
freqPlot.publications2
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are (in descending order): 
 1. "Lab" 
 2. "Field Natural" 
 3. "Field Artificial"
 4. "Undeterminable"
As time increases, the number of lab locations increases dramatically. Natural fields spike, then decrease.
Artificial labs are stable, as are undeterminable fields.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r bargraph.plot}
# Create and save plot
barGraph.endpoints <- ggplot(Neonics, aes(x = Endpoint))+
  # Set bar chart type
  geom_bar() +
  # Set title and subtitle
  ggtitle("Endpoint Classification Count", 
          "Source: National Ecological Observatory Network") +
  # Change x- and y-axis labels
  labs(x = "Endpoint", y = "Endpoint Count") +
  # Adjust plot text to make more legible
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  # Add bar counts to top of individual bars
  geom_text(size = 2.8, aes(label=..count..),
            stat='count',
            position=position_dodge(0.5),
            vjust=-0.27)

# Call bar graph
barGraph.endpoints
```

> Answer: The two most common endpoints are:
  1. L0EL (Lowest-observable-effect-level): lowest dose (concentration)
producing effects that were significantly different (as reported
by authors) from responses of controls
  2. NOEL (No-observable-effect-level): highest dose (concentration)
producing effects not significantly different from responses of
controls according to author's reported statistical test

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r litter.collectDate}
class(Litter$collectDate)

Litter$collectDate <- as.Date(Litter$collectDate, "%Y-%m-%d")
Litter$collectDate
```

```{r litter.class}
class(Litter$collectDate)

unique(Litter$collectDate)
```

> Answer: Litter was sampled on two days: 08/02/18 and 08/30/18

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r unique.litter}
# Retrieve unique values in plot ID field
unique(Litter$plotID)
# Retrieve summary stats of plot ID field
summary(Litter$plotID)

# sum number of unique values in plot ID field
sum(summary(unique(Litter$plotID)))
```

> Answer: 12 plots were sampled at Niwot Ridge.
The information obtained from unique() returns the specific plot IDs, without number of occurrences.
The summary() function returns each plot ID as well as the number of occurrences.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r bargraph.functionalgroup}
# Create and save bar graph
barGraph.functionalGroup <- ggplot(Litter, aes(x = functionalGroup)) +
  # Set geometry bar graph type
  geom_bar() +
  # Set title and subtitle
  ggtitle("Litter Functional Group Count", 
          "Source: National Ecological Observatory Network") +
  # Set x- and y-axis labels
  labs(x = "Functional Group", y = "Functional Group Count") +
  # Adjust label orientation
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  # Add bar counts, and adjust label position
  geom_text(size = 3, aes(label=..count..),
            stat='count',
            position=position_dodge(0.5),
            vjust=-0.27)

# Call bar graph
barGraph.functionalGroup
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r box.violin.plot}
# Create boxplot
boxPlot.functionalGroup <- ggplot(Litter, aes(x = dryMass, y = functionalGroup)) +
  # Set boxplot type
  geom_boxplot() +
  # Set title and subtitle
  ggtitle("Mass Distributions by Functional Group", 
          "Source: National Ecological Observatory Network") +
  # Set x- and y-axis labels
  labs(x = "Mass", y = "Functional Group Count") +
  theme_minimal()

# Call box plot
boxPlot.functionalGroup

violinPlot.functionalGroup <- ggplot(Litter, aes(x = dryMass, y = functionalGroup)) +
  geom_violin() +
  # Set title and subtitle
  ggtitle("Mass Distributions by Functional Group", 
          "Source: National Ecological Observatory Network") +
  # Set x- and y-axis labels
  labs(x = "Mass", y = "Functional Group Count") +
  theme_minimal()

# Call violin plot
violinPlot.functionalGroup
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is more effective in this case because:
1. It provides meaningful summary statistics, such as median, quartiles, etc.
2. Shows outliers
The violin plot does not show any density around the the center lines, possibly 
because the mass distributions are so small.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles (on average)