---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Keanu Valibia"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

\tableofcontents
\listoffigures

## Overview

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup1, include=FALSE}

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=45), tidy=TRUE)

```


```{r setup2, echo=TRUE, results='hide', message=FALSE}
#1

library(tidyverse)
library(agricolae)
library(lubridate)
library(here)
library(ggplot2)

getwd()
here()

ntl.lter.raw <- read.csv(here("~/R/R Projects/EDA_Spring2024/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))
ntl.lter.raw$sampledate <- as.Date(ntl.lter.raw$sampledate, format = '%m/%d/%y')

#2

myTheme <- theme_classic(base_size = 11) +
  theme(axis.text = element_text(color = "black"),
        axis.line = element_line(color = "#320E3B"),
        panel.background = element_rect(fill ="#EDE6E3"),  
        panel.grid.major = element_line(color = "#36382E",  
                                        linetype = "dotted"),  
        plot.title = element_text(size = 15),  
        axis.title.x = element_text(size = 13),  
        axis.title.y = element_text(size = 13),
        legend.position = "right")

theme_set(myTheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: mu = 0 (There is not enough reason to reject the null hypothesis: mean temperatures recorded in July do not vary across all lakes)
Ha: mu != 0 (There is reason to reject the null hypothesis: mean temperatures recorded in July do not vary across all lakes)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.


```{r scatterplot, fig.cap = "Depth-To-Temperature Analysis", warning=FALSE, message=FALSE, fig.align='center'}
#4

ntl.lter.wrangled <- ntl.lter.raw %>%
  filter(month(sampledate) %in% 5) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5

tempByDepth <- 
  ggplot(ntl.lter.wrangled, aes(x = depth, y = temperature_C)) +  
  geom_smooth(method = "lm", color = "#320E3B") +  
  labs(title = "Depth-to-Temperature Analysis",  
       caption= "Source: Source: North Temperate Lakes Long-Term Ecological Research Station") +  
  xlab("Depth") +  
  ylab("Temperature (Celsius)") +  
  ylim(0, 35) +  
  geom_point()

tempByDepth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatterplot suggests a negative relationship between Temperature and Depth. However, there is a certain point in which temperatures flatten (approximately 5 degrees). Perhaps at temperatures closer to 0 (since it's freezing temperature of water), there seems to be a logical floor in which we can measure the temperature of water before it solidifies.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

summary(ntl.lter.wrangled)

regression <- lm(data = subset(ntl.lter.raw, month(sampledate) %in% 5), 
                   temperature_C ~ depth)

summary(regression)

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(regression)
par(mfrow = c(1,1))

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: R-squared is 0.638, which indicates that the model explains about 63.8% of the variability between changes in dpeth and temperature. The DF is 4079. The linear regression model results in a p-value of 2.2e-16, which is statistically significant at the 0% level. Temperature is predeicted to drop by -1.16 degrees for every 1m increase in change in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

#Choose a model by AIC in a Stepwise Algorithm

AIC <- lm(data = ntl.lter.wrangled, temperature_C ~ depth + year4 + daynum)
AIC

step(AIC)

#10

AICmodel <- lm(data = ntl.lter.wrangled, temperature_C ~ depth + year4 + daynum)
summary(AICmodel)

AICmodel

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables are the same as the original input: year, depth, and day number. The R-squared is 0.6476, which means that 64.76% of the variation is explained by the model. The p-values are the same, however the AIC model has a slight increase of the Adjusted R-Squared which indicates that the AIC model slightly explains the variability more than the linear regression model. Additionally, the coefficients for depth only decreases by ~0.001 from a linear regression model to an AIC model. These differences are only marginal.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

lake.anova <- aov(data = ntl.lter.wrangled, temperature_C ~ lakename)

summary(lake.anova)

lake.lmRegression <- lm(data = ntl.lter.wrangled, temperature_C ~ lakename)
summary(lake.lmRegression)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperatures among the lakes. The overall model p-value is significant at the 0% level with a p-value of 2.61e-15. When examined at the coefficient-level, all lakes have a significant difference all at the 0% level.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.cap = "Depth-To-Temperature Analysis by Lake", warning=FALSE, message=FALSE, fig.align='center'}
#14.

lake.tempByDepth.plot <- ggplot(ntl.lter.wrangled, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_smooth(method = "lm", se = FALSE, color = "#320E3B") +
  labs(title = "Depth-to-Temperature Analysis by Lake",
       caption= "Source: Source: North Temperate Lakes Long-Term Ecological Research Station",
       color = "Lake Name") +
  xlab("Depth") +
  ylab("Temperature (Celsius)") +
  ylim(0, 35) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values=c("#642915", "#963e20", "#c7522a", "#e5c185", "#fbf2c4", "#74a892", "#008585", "#006464", "#004343"))

lake.tempByDepth.plot

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

lake.tukey <- TukeyHSD(lake.anova)
lake.tukey

tukey.lakeGroups <- HSD.test(lake.anova, "lakename", group = TRUE)
tukey.lakeGroups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the closest statistically significant mean temperatures at the 95% Confidence Level are Peter Lake and East Long Lake (Diff: 0.86 at p-value: 0.042). However, Ward Lake, Crampton Lake, and Paul Lake are all grouped similarly to Peter Lake. Central Lake has a mean temperature that is statistically different from all other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We may use a two-sided t-test to compare the means of the two lakes, using Peter and Paul as the categorical variable.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

ntl.lter.wrangled2 <- ntl.lter.wrangled %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

ntl.twoSample <- t.test(temperature_C ~ lakename, ntl.lter.wrangled2)
ntl.twoSample

```

>Answer: The mean tempreatures for the lakes seem to be equal, but are not signficant at the 95% Confidence Level (p-value = 0.9524). There is not enough evidence to reject the null hypothesis (mu = 0). These results match the answer I received in part 16, which indicates a mean temperature difference of ~0.03 that is not statistically significant at the 95% Confidence Level.